Analysis of Programming Students' Prompts to Identify Patterns of Critical Thinking in Interactions with Generative AI \\ {\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and should not be used} \thanks{Identify applicable funding agency here. If none, delete this.}

### 2024-08-12 paper/paper.tex
update

\author{\IEEEauthorblockN{1\textsuperscript{st} Rodrigo Prestes Machado}
\IEEEauthorblockA{\textit{dept. Informática} \\
\textit{Instituto Federal do Rio Grande do Sul}\\
Porto Alegre, Brazil \\
0000-0003-0428-6387}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{5\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{6\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
}

@@

According \cite{Kwan24} further research about the use of AI tools in
education is needed to understand how these tools can be used to improve
students' learning experience. In particular, the study suggested that future
research on students' study habits.

Problems

The literature suggests a concern regarding the impact of ChatGPT on students'
critical thinking abilities. Vargas \cite{Vargas2023} indicated that
ChatGPT use might lead to an overreliance on the tool, potentially diminishing
students' critical thinking skills. Similarly, Cai \cite{cai23} identified
overdependence and reduced intellectual engagement as significant drawbacks of
ChatGPT-supported learning. Chan \cite{chan23} also noted that this
reliance could result in a decrease in critical thinking, as students might make
decisions based solely on the information provided by ChatGPT. Collectively,
these studies highlight a common theme of reduced critical engagement and
overreliance on ChatGPT in educational contexts.

@@

@@

@@

@@

@@

The paper \cite{Kwan24} examined students from different demographic groups
using AI tools like ChatGPT, comparing perceptions and impacts
among these groups. The results showed that AI tools had a positive impact on
academic performance, especially for students with less prior experience.

@@

Finally, some students asserted that the use of ChatGPT enhanced the depth of
their thinking and understanding \cite{chan23}.

>>>>

\author{\IEEEauthorblockN{1\textsuperscript{st} Rodrigo Prestes Machado}
\IEEEauthorblockA{\textit{dept. Informática} \\
\textit{Instituto Federal do Rio Grande do Sul}\\
Porto Alegre, Brazil \\
0000-0003-0428-6387}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Carlos Alario Hoyos}
\IEEEauthorblockA{\textit{Dep. Ingeniería Telemática} \\
\textit{Universidad Carlos III de Madrid}\\
Madrid, Spain \\
0000-0002-3082-0814}
\and
\IEEEauthorblockN{3\textsuperscript{rd} Carlos Delgado Kloos}
\IEEEauthorblockA{\textit{Dep. Ingeniería Telemática} \\
\textit{Universidad Carlos III de Madrid}\\
Madrid, Spain \\
0000-0003-4093-3705}
}

@@

Recent advancements in Generative Artificial Intelligence (GenAI) have opened
up new possibilities in education. In programming courses, students can utilize
GenAI tools to improve their understanding, receiving assistance, feedback, and
detailed explanations.

The study by Chan and Hu et al. \cite{chan23} revealed that both undergraduate
and postgraduate students exhibit positive attitudes toward the use of GenAI
in teaching and learning, noting that it enhances the depth of their thinking
and understanding. Additionally, a systematic review by Lo, Hew, and Jong
\cite{Lo24} demonstrated that students could effectively learn from ChatGPT,
leading to improved understanding and academic achievement. Cai et al.
\cite{cai23} further observed that ChatGPT enables students to regulate their
learning pace, promoting self-regulation.

In the other hand, researchers have concerns about the impact of these
tools on students. The systematic review of Vargas-Murillo et al.
\cite{Murillo23} indicated that ChatGPT use might lead to an overreliance on the
tool. Chan et al. \cite{chan23} also noted that this reliance could result in a
decrease in critical thinking, as students might make decisions based solely on
the information provided by ChatGPT.

The student's confidence is warranted, as shown by Puryear and Sprint
\cite{Puryear22}, who demonstrated that GitHub Copilot can generate solutions
for student assignments with accuracy rates ranging from 68\% to 95\%. However,
this raises concerns that students might rely too heavily on GenAI tools,
potentially neglecting a deeper understanding of the underlying concepts.
Cai et al. \cite{cai23} further identified overdependence and reduced
intellectual engagement as significant drawbacks of using ChatGPT in learning.

Regardless of teachers' preferences or beliefs, preliminary surveys by Dickey,
Bejarano, and Garg \cite{Dickey24} revealed that at least 54.5\% of students are
already using Generative AI (GenAI) for homework. This highlights the need to
understand how students are interacting with these tools and how they can be
used to enhance learning. Furthermore, a review by Lo, Hew, and Jong \cite{Lo24}
emphasized the necessity for extended studies and objective measures to gather
more robust evidence on the use of GenAI tools in education.

@@

To deal with this issue, some researchers have proposed different pedagogical
strategies. Denny et al. \cite{Denny24} introduced the concept of \textit{Prompt
Problems}, where students solve programming exercises by formulating natural
language prompts. Prasad and Aamod \cite{Prasad24} proposed a self-regulated
learning framework using GenAI to solve programming problems. Lauren and Watta
\cite{Lauren23} explored the integration of generative AI with evidence-based
learning strategies in computer science and engineering courses.

@@

Effective strategies are crucial to determinate the best approaches and training
educators to use GenAI tools effectively can be time-consuming and
resource-intensive. Therefore, understanding how students interact with these
tools is essential. By analyzing these interaction patterns, we can assess the
impact of GenAI tools on students' learning outcomes.
This knowledge can then be used to refine the design of GenAI tools, ensuring
they better support pedagogical strategies and enhance overall learning
experiences.

@@

It is important to consider how GenAI tools affect students across different
demographic groups, knowledge areas, and levels of programming experience.
Understanding these variations can provide insights into the impact of GenAI
tools on their learning outcomes, helping to tailor educational approaches that
address the specific needs of diverse student populations.

@@

This study aims to analyze the interactions between students and CharlieBot,
an ChatGPT-based bot developed by UC3M that leverages Retrieval-Augmented
Generation (RAG) to enhance its contextual understanding of Java programming.

@@

The paper \cite{Lo24} From the cognitive perspective, students were able to
learn from ChatGPT, which increased their understanding and achievement.
However, concerns were raised that the growing use of AI tools might lead to
a decline in critical thinking among students. Teachers and stakeholders should
continue to investigate pedagogical approaches that leverage ChatGPT to enhance
students' understanding and critical thinking. For example, teachers can
instruct students to fact-check and validate information produced by ChatGPT.

examined students from different demographic groups
using AI tools like ChatGPT, comparing perceptions and impacts
among these groups. The results showed that AI tools had a positive impact on
academic performance, especially for students with less prior experience.

@@
